Altering Backward Pass Gradients improves Convergence

### 2022-09-16 paper.tex
Modifications 3

Backpropagation is traditionally used to train deep neural networks
\cite{lillicrap2020backpropagation}. Gradients are computed using basic calculus
principles to adjust the weights during backpropagation \cite{lecun1988theoretical}.
Alternatives to traditional gradients has rarely been studied in the literature
hitherto. In normal training procedures, gradients are computed immediately in the
backward pass utilizing the values obtained in the forward pass. This makes the backward
pass coupled with the forward pass. However, decoupling the backward pass from the
forward pass by modifying the gradients to improve training efficiency and final
convergent accuracy has hardly been explored. In this paper we explore the landscape of
decoupling the forward pass from the backward pass by altering the gradients and
subsequently updating the network's parameters with the modified gradients. There are
several ways to achieve gradient modification in the backward pass. We discuss a few
techniques in Fig. \ref{fig:gradient_altering}. One could modify the gradients at either
\textbf{(I)} multiple points in the backward computation graph or \textbf{(II)} at a
point very early in the backward graph and allow this changed gradient to also affect
the rest of the network's gradients as the it backpropagates through the backward
computation graph.

@@

\textbf{Type II:} Gradient modification at a point very early in the backward graph: In
this type of modification, the gradient is altered at a very early stage in the backward
computation graph and then all subsequent gradients are generated using the values
obtained with the modified gradients (Fig. \ref{fig:gradient_altering}c). Because of the
chain rule, network parameters whose gradients are connected to the point of alteration
the computation graph also gets subsequently altered. It can be described as:

\begin{align}
\hspace{4cm}
W_D^{t+1} &= W_D^t - \lambda\ f(\nabla_{W_D}(L)) & \\
W_i^{t+1} &= W_i^t - \lambda\ \nabla_{W_i}(L) &\hspace{-3cm} i=D-1,\ldots,1
\end{align}

where first the gradient $\nabla_{W_D}(L)$ is transformed using the transformation
function `$f$' and then this transformed gradient is propagated through the rest of the
backward graph and all other gradient vectors are computed as is. Because of the early
injection of the transformed gradients, all other gradient vectors that are connected to
the transformed gradient through the chain rule, gets subsequently altered.

@@

PowerGrad Transform modifies the gradients at the softmax layer, which is the earliest
part of most convolutional neural network's computation graph. By changing the gradient
at this stage, all other gradients for all other parameters in the network such as the
linear layer, batchnorm statistics and convolutional layer's filter parameters are
affected. We analyze and explore the behaviour of such alteration both mathematically
and through empirical experiments conducted on networks without batch-normalization.
Experimentally, we see that PGT improved training and test accuracies for both
normalized and non-normalized networks. PGT improves a network's learning capacity and
results in a better fit of the training data.

@@

\item We introduce PowerGrad Transform, which decouples the backward and the forward
passes of neural network training and enables a considerably better fit to the dataset,
as assessed by both training and test accuracy measures. PGT is a gradient enhancement
method that alters the gradients in the backward pass before the update step leading to
significant boost in network's predictive performance.

@@

Adaptive Gradient Clipping (AGC): Adaptive Gradient Clipping \cite{brock2021high} is
developed to further enhance backward pass gradients than what is performed by GC. It
takes into account the fact that the ratio of the gradient norm to the weight norm can
provide an indication of the expected change in a single step of optimization. Here the
normalized gradient is clipped by $\lambda$, which means that large weights can have
large gradients and still the normalized gradient can be within $\lambda$. If it is not
within the threshold, then their gradient is normalized by the ratio of norm of the
gradient and norm of the weight in order to avoid gradient explosion. If $W^l$ is the
weight matrix of the $l^\textit{th}$ layer, and $G^l$ is its gradient in the backward
pass, then the following equation is used to modify the gradient of the $i^\textit{th}$
filter before its update. In most cases, AGC is used for training non-BN networks. Our
experiments demonstrate that PGT outperforms AGC, when used independently. When used
together, it can further improve a network's performance.

@@

PowerGrad Transform works in a much different way. PGT does not smooth the ground truth
labels, rather it modifies the gradient by smoothing the predicted probability vector.
As we show in section \ref{sec:Math}, our proposed transformation has the opposite
effect of label smoothing. From our studies, we show that PGT in fact leads to larger
generalization gap between the training and test dataset. Coupled with PGT, label
smoothing can narrow the generalization gap.

@@

Two other methods that also changes the gradients in the backward pass are \textbf{i)}
Gradient Clipping \cite{pascanu2013difficulty} \textbf{ii)} Adaptive Gradient Clipping
\cite{brock2021high}. We introduce these two methods before delving into the inner
workings of PowerGrad Transform.

The hyperparameter $\lambda$ of AGC is less sensitive to changes in batch size and depth
as observed in \cite{brock2021high}. However, when applied to vanilla residual networks
(ResNet-18), the performance is still afar from normalized variants as demonstrated by
our experiments.

@@

We introduced PowerGrad Transform, which decouples the backward and forward passes of
neural network training and enables a significantly better fit to the dataset, as
measured by training and test accuracy metrics. We show the application of PowerGrad
Transform, a simple yet powerful technique for modifying the gradient flow behaviour. We
investigate experimentally the degenerate behavior of non-BN networks, which is
frequently observed during the standard training approach, especially with higher batch
sizes. With PGT, gradient behavior is enhanced and the likelihood of weights attaining
degenerate states is diminished. We provide a theoretical analysis of the PowerGrad
transformation. With the use of different network topologies and datasets, we are able
to show the potential of PGT and explore its impacts from a theoretical standpoint. We
provide comprehensive results on a number of models (non-BN and BN ResNets, SE-ResNets)
using the ImageNet dataset. PGT helps the network to improve its learning capabilities
by locating a more optimum convergence point. In addition, we undertake an ablation
research and compare its effects to those of regularization methods such label
smoothing.

>>>>

Backpropagation is traditionally used to train deep neural networks
\cite{lillicrap2020backpropagation}. Gradients are computed using basic calculus
principles to adjust the weights during backpropagation \cite{lecun1988theoretical}.
Alternatives to traditional gradients has rarely been studied in the literature
hitherto. In normal training procedures, gradients are computed immediately in the
backward pass utilizing the values obtained in the forward pass. This makes the backward
pass coupled with the forward pass. However, decoupling the backward pass from the
forward pass by modifying the gradients to improve training efficiency and final
convergent accuracy has hardly been explored. In this paper we explore the landscape of
decoupling the forward pass from the backward pass by altering the gradients and
subsequently updating the network's parameters with the modified gradients. There are
several ways to achieve gradient modification in the backward pass. We discuss a few
techniques in Fig. \ref{fig:gradient_altering}. One could modify the gradients at either
\textbf{(I)} multiple points in the backward computation graph or \textbf{(II)} at a
point very early in the backward graph and allow this changed gradient to also affect
the rest of the network's gradients as it backpropagates through the backward
computation graph.

@@

\textbf{Type II:} Gradient modification at a point very early in the backward graph: In
this type of modification, the gradient is altered at a very early stage in the backward
computation graph and then all subsequent gradients are generated using the values
obtained with the modified gradients (Fig. \ref{fig:gradient_altering}c). Because of the
chain rule, network parameters whose gradients are connected to the point of alteration
in the computation graph also gets subsequently altered. It can be described as:

\begin{align}
\hspace{4cm}
W_D^{t+1} &= W_D^t - \lambda\ f(\nabla_{W_D}(L)) & \\
W_i^{t+1} &= W_i^t - \lambda\ \nabla_{W_i}(L)^{*} &\hspace{-3cm} i=D-1,\ldots,1
\end{align}

where first the gradient $\nabla_{W_D}(L)$ is transformed using the transformation
function `$f$' and then this transformed gradient is propagated through the rest of the
backward graph and all other gradient vectors $\nabla_{W_i}(L)^{*}$ are computed as is.
Because of the early injection of the transformed gradients, all other gradient vectors
that are connected to the transformed gradient through the chain rule, gets subsequently
altered.

@@

PowerGrad Transform modifies the gradients at the softmax layer, which is the earliest
part of most convolutional neural network's backward computation graph. By changing the
gradient at this stage, all other gradients for all other parameters in the network such
as the linear layer, batchnorm statistics and convolutional layer's filter parameters
are affected. We analyze and explore the behaviour of such alteration both
mathematically and through empirical experiments conducted on networks without
batch-normalization. Experimentally, we see that PGT improved training and test
accuracies for both normalized and non-normalized networks. PGT improves a network's
learning capacity and results in a better fit of the training data.

@@

\item We introduce PowerGrad Transform, which decouples the backward and the forward
passes of neural network training and enables a considerably better fit to the dataset,
as assessed by both training and test accuracy measures. PGT is a performance
enhancement method that alters the gradients in the backward pass before the update step
leading to significant boost in network's predictive performance.

@@

Adaptive Gradient Clipping (AGC): Adaptive Gradient Clipping \cite{brock2021high} is
developed to further enhance backward pass gradients than what is performed by GC. It
takes into account the fact that the ratio of the gradient norm to the weight norm can
provide an indication of the expected change in a single step of optimization. Here the
normalized gradient is clipped by $\lambda$, which means that large weights can have
large gradients and still the normalized gradient can be within $\lambda$. If it is not
within the threshold, then their gradient is normalized by the ratio of norm of the
gradient and norm of the weight in order to avoid gradient explosion. If $W^l$ is the
weight matrix of the $l^\textit{th}$ layer, and $G^l$ is its gradient in the backward
pass, then the following equation is used to modify the gradient of the $i^\textit{th}$
filter before its update:

The hyperparameter $\lambda$ of AGC is less sensitive to changes in batch size and depth
as observed in \cite{brock2021high}. However, when applied to vanilla residual networks
(ResNet-18), the performance is still afar from normalized variants as demonstrated by
our experiments. In most cases, AGC is used for training networks without
batch-normalization. Our experiments demonstrate that PGT outperforms AGC, when used
independently. When used together, it can further improve a network's performance
(section \ref{sec:Empi}).

@@

PowerGrad Transform works in a much different way. PGT does not smooth the ground truth
labels, rather it modifies the gradient by smoothing the predicted probability vector.
As we show in section \ref{sec:Math}, our proposed transformation has the opposite
effect of label smoothing. From our studies, we show that PGT in fact leads to larger
generalization gap between the training and test dataset, although it does lead to an
increase in both training and test accuracies at convergence. Coupled with PGT, label
smoothing can narrow the generalization gap, while retaining the benefits of PGT.

@@

@@

We introduced PowerGrad Transform, which decouples the backward and forward passes of
neural network training and enables a significantly better fit to the dataset, as
measured by training and test accuracy metrics. We show the application of PowerGrad
Transform, a simple yet powerful technique for modifying the gradient flow behaviour. We
investigate experimentally the degenerate behavior of non-BN networks, which is
frequently observed during the standard training approach, especially with higher batch
sizes. With PGT, gradient behavior is enhanced and the likelihood of weights attaining
degenerate states is diminished. We provide a theoretical analysis of the PowerGrad
transformation. With the use of different network topologies and datasets, we are able
to show the potential of PGT and explore its impacts from an empirical standpoint. We
provide comprehensive results on a number of models (non-BN and BN ResNets, SE-ResNets)
using the ImageNet dataset. PGT helps the network to improve its learning capabilities
by locating a more optimum convergence point. In addition, we undertake an ablation
research and compare its effects to those of regularization methods such label
smoothing.
